---
title: "calanusthreshold"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Provides R tools for producing threshold prey models for Calanus.

## Requirements

 + [R v4.1+](https://www.r-project.org/)

 + [rlang](https://CRAN.R-project.org/package=rlang)
 
 + [dplyr](https://CRAN.R-project.org/package=dplyr)
 
 + [readr](https://CRAN.R-project.org/package=readr)
 
 + [recipes](https://CRAN.R-project.org/package=recipes)
 
 + [rsample](https://CRAN.R-project.org/package=rsample)
 
 + [parsnip](https://CRAN.R-project.org/package=parsnip)
 
 + [yardstick](https://CRAN.R-project.org/package=yardstick)
 

## Installation

```
remotes::install_github("BigelowLab/calanusthreshold")
```

**Convention**

> We prepend the package namespace to function **not** exposed by the `calanusthreshold` package.  Bare function names (no package prepended) are used for base R functions and for `calanusthreshold` functions.

## The input data - observations

We provide an anonymized example dataset.

```{r read_example, message = FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(calanusthreshold)
  library(tidymodels)
})
x <- read_dataset()
dplyr::glimpse(x)
```

If you have your own dataset just provide the filename and path, but be sure to see `?read_dataset` dcoumentation.  Note that the variables with species names are abundance measurements by life stage.

```
x <- read_dataset("/path/to/input/data.csv")
```

We are modeling patches of the species which we define by excess abundance over user specified thresholds. For this example we define just one  threshold...

```{r}
THRESHOLD <- 10000
```

For the purpose of illustration, we lump all stages of a given species together. 

```{r}
Species <- calanusthreshold::known_species()
abund <- lapply(Species,
                        function(s) {
                          dplyr::tibble(species = s,
                            abund = calanusthreshold::lump_vars(x, vars = s,
                                                      selector = dplyr::starts_with) |>
                                    dplyr::pull("lumped"))
                        })|>
                  dplyr::bind_rows() |>
  dplyr::mutate(patch = factor(findInterval(abund, THRESHOLD),
                               levels = seq(from = 0, to = length(THRESHOLD)),
                               labels = c("sparse", THRESHOLD))) |>
  dplyr::group_by(species) |>
  dplyr::glimpse()
```

We can plot the raw abundance by species group, note that we drop records where abundance is less than 0 for the sake of clarity. Note that *C glacials* generally has low patchiness when the threshold is held high.

```{r}
ggplot2::ggplot(abund %>% dplyr::filter(abund > 0), ggplot2::aes(species, abund)) + 
  ggplot2::geom_violin() + 
  ggplot2::scale_y_log10() +
  ggplot2::labs(title = "Filtered input raw abundances (>0) by species",
       x = "Species", y = "Abundance (log10)") +
  ggplot2::geom_hline(yintercept = THRESHOLD, 
             linetype = 'dashed',
             color = 'grey')
```

## The model - random forest

The [tidymodels](https://www.tmwr.org/) attempts to provide a uniform programmatic interface for myriad model functions/packages in R, including [random forest](https://en.wikipedia.org/wiki/Random_forest) models.  The idea is to provide consistent argument interfaces, "engine" selection, and uniform model assessment metrics. 

We are going to model just `Calanus Finmarchicus`.  First we prep the dataset - dropping non-essential variables (columns), log=scaling some variables and by aggregating (lumping) the various life-stages. The `prep_dataset()` function includes a 'patch' label indicating that the abundance for that observation exceeded the threshold. We then split the data into a user specified proportion of training and testing groups.

```{r}
x_split <- prep_dataset(x,
                        lump_var = "Calanus finmarchicus",
                        threshold = THRESHOLD) |>
 dplyr::relocate(patch, .before = 1) |>
 rsample::initial_split(prop = 0.6)
x_split
```

Next we build up a series of 'recipe' steps that prepare the data. The `prep` steps implements all of the other steps, including removal highly correlated variable (like roughness).

```{r}
x_recipe <- rsample::training(x_split) |>
    recipes::recipe(patch ~.) |>
    recipes::step_corr(recipes::all_predictors()) |>
    recipes::step_center(recipes::all_predictors(), -recipes::all_outcomes()) |>
    recipes::step_scale(recipes::all_predictors(), -recipes::all_outcomes()) |>
    recipes::prep()
x_recipe
```

Now we can 'bake' the training and testing datasets.

```{r}
x_testing <- x_recipe |>
    recipes::bake(rsample::testing(x_split)) 
  
  x_training <- x_recipe |>
    recipes::bake(rsample::training(x_split))
```

Now we build and fit a model.  Note the we are accepting the defaults arguments for the [ranger::ranger()](https://www.rdocumentation.org/packages/ranger/versions/0.13.1/topics/ranger) function.

```{r}
x_model <- parsnip::rand_forest(mode = "classification") |>
      parsnip::set_engine("ranger") |>
      parsnip::fit(patch ~ ., data = x_training)
x_model
```

##  The prediction

We can now predict the outcome if we pass in another dataset, in the case the testing data we reserved as 40% of the original.  We'll bind the prediction to the testing data and reorganize for clarity. In the output `patch` is out original (aka 'Truth') while `.pred_class` is the predicted class (aka 'Predicted')

```{r}
x_pred <-  predict(x_model, x_testing) |>
    dplyr::bind_cols(x_testing) |>
    dplyr::relocate(patch, .before = 1)
dplyr::glimpse(x_pred)
```

We can compute and examine a confusion matrix.

```{r}
x_cm <- yardstick::conf_mat(x_pred, patch, .pred_class)
x_cm
```

Or present graphically...

```{r}
heat_map(x_cm, title = "Calanus finmarchicus")
```

Similar the the ease of computing a predicted class value, we can also produce predicted class probabilities. Here we 

```{r}
x_prob <-  predict(x_model, x_testing, type = "prob") |>
  dplyr::bind_cols(x_testing) |>
  dplyr::relocate(patch, .before = 1) |>
  dplyr::glimpse()
```

From the probabilities for each class we can compute ACU on the ROC.

```{r}
x_auc <- yardstick::roc_auc(x_prob, patch, .pred_1, event_level = 'second') |>
  dplyr::glimpse()
```